---
title: "Data cleaning"
author: "Lauren Harris"
date: "7/1/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(tidyverse)
library(here)
library(lubridate)
```

Import data
```{r}
spp_data_0 <- read.csv(here("data", "input", "living_species_present_1.csv"))

plot_data_0 <- read.csv(here("data", "input", "Form_1_0.csv"))
```

Clean up columns labeling for datasets
```{r}
spp_data <- spp_data_0 %>% # reduce and rename columns 
  select(global_id_b = GlobalID, 
         spp_name = Species.name, 
         spp_name_other = Specify.other.,
         relative_cover = X..cover.of,
         global_id_a = ParentGlobalID, 
         date = CreationDate) %>% 
  mutate(date = mdy_hms(date)) # convert date to date class

plot_data <- plot_data_0 %>% # reduce and rename columns
  select(global_id_a = GlobalID, 
         plot = Plot, 
         relative_plant = X..Living.Plant.Cover, 
         relative_thatch = X..thatch.cover, 
         relative_bare = X..Bare.Ground, 
         percent_native = X..Living.natives, 
         percent_nonnative = X..Living.non.natives, 
         thatch_depth_in = Thatch.depth..inches.,
         burrow_count = Number.of.animal.burrows,
         lat, 
         lon, 
         date = Date.and.Time) %>% 
  mutate(date = mdy_hms(date)) # convert date to date class
```

Add in plot labels using global IDs

The global ID in the plot data matches the parent global ID in the spp data
```{r}
# reference list for plot names to global IDs
plot_name_list <- plot_data %>% 
  select(global_id_a, plot) %>% 
  distinct()

# join the ref and the spp list to put plot names to species
spp_data <- left_join(x = spp_data, y = plot_name_list, by = "global_id_a")
```

Basic summary statistics
```{r}
plot_richness <- spp_data %>% 
  group_by(month(date), year(date), plot) %>% 
  summarize(spp_richness = n()) # still want to rename date columns

# plot broken

plot_richness %>%
  tail(10) %>%
  ggplot(aes(x=year, y=value)) +
    geom_line() +
    geom_point()
```

To do list: 
- create csv of plots and whether they are control or experimental
- put "other" species in main species column
- create a csv with plant nativity
- explore plant family/higher taxonomy effects


Plots to create:
- biodiversity
- richness
